Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-03 index.qmd
section:issues,CaseV; process:write

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone's theory deals with pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp. 267]. The theory proposes that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their underlying perception of the stimulus' trait level. According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (or the mean), referred to as the *modal discriminal process*, indicates the stimulus' position on the trait continuum, and its dispersion, known as the *discriminal dispersion*, reflects the variability in the perceived trait level of the stimulus.

@@

Indeed, the theory assumes that the observed dichotomous outcome arises from the distribution of the difference between the underlying discriminal processes of the stimuli, known as the *discriminal difference*. Since the individual discriminal processes follow a Normal distribution, their difference also follows a Normal distribution [@Andrich_1978]. The mode (or the mean) of this distribution, representing the (average) relative separation, is given by the difference between the modal discriminal processes of the stimuli $S_{BA} = S_{B} - S_{A}$. Meanwhile, the dispersion of the distribution, reflecting the variability in the relative separation, is calculated as $\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}$. Here, $\sigma_{B}$ and $\sigma_{A}$ are the previously defined discriminal dispersions, while $\rho$ measures the correlation between the discriminal processes of the stimuli. This correlation quantifies the dependence of the judge's perception of the trait in one stimulus on his perception of the same trait in the other.

@@

<!-- 6. example of the effects of correlation on the discriminal difference -->
@fig-correlation illustrates how varying correlations influence the distribution of the discriminal difference for the two texts depicted in @fig-discriminal_process. Since the texts differ in quality, higher correlations increase the likelihood that the discriminal difference distinctly favors text B over text A. This is evident from the larger proportion of the area under the curve, $P(B > A)$, that lies above zero (shaded gray area in @fig-correlation). Conversely, it is not hard to infer that if the texts had similar or identical quality levels, higher correlations would likely reduce the chance that the discriminal difference distinctly favors one text over the other. This probability reduction occurs because the distribution of the discriminal difference would become more narrowly centered around zero (not illustrated).

@@

<!-- 1. Thurstone's Cases and their simplifying assumptions  -->
The previous section outlines the general form of Thurstone's theory, which applies to a CJ design where a single judge evaluates multiple stimuli. For the practical application of the theory, Thurstone developed four additional cases derived from this general form, where each successive case incorporates additional simplifying assumptions. Case I represents the general form of the theory. Case II extends this by allowing multiple judges to make comparisons rather than restricting them to a single judge. Case III introduces the assumption of zero correlation between stimuli. Case IV builds on this by assuming that the stimuli have similar dispersions. Finally, Case V replaces this assumption with the condition that the stimuli have equal discriminal dispersions. @tbl-thurstone_cases summarizes these cases and their assumptions. For a detailed discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248-253].

@@

<!-- 5. The first scenario where the assumption does not hold -->
In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between the stimuli due to certain judges' biases that resist cancellation. Research on text quality suggests that when judges evaluating these traits, they often rely on various intricate characteristics of the stimuli [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. These additional relevant characteristics, which are unlikely to be equally weighted or opposing, can unevenly influence judges' perceptions, creating biases in their judgments and, consequently, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp. 346]. For example, this could occur when a judge assessing the argumentative quality of a text places disproportionate emphasis on grammatical accuracy, ultimately favoring texts with fewer errors but weaker arguments. While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the idea that the factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 7. But what are the challenges when these assumptions are violated? and how to solve them? -->
In any case, the psychometric and statistical literature underscores the importance of incorporating additional relevant traits and clustering (grouping) structures into the modeling of CJ data to address potential dependencies between stimuli. Oversights in these areas can lead to misleading conclusions about the trait's reliability and ultimately compromise its validity [@Perron_et_al_2015, pp. 2]. Neglecting additional traits relevant to the stimuli or the trait of interest, such as judges' biases, introduces a dimensional mismatch in the modeling process of the CJ data. This mismatch often results in an overestimation of the trait's reliability [@Hoyle_et_al_2023, pp. 341] or, worse, introduces biases into the estimated trait [@Ackerman_1989]. Therefore, these traits must be incorporated into the modeling of CJ data, as demonstrated by @Andrich_1978 and @Wainer_et_al_1978. Similarly, ignoring hierarchical (grouping) structures reduces the accuracy of model parameter estimates and further exacerbates the overestimation of reliability [@Hoyle_et_al_2023, pp. 482]. To address this, any additional hierarchical structure relevant to the stimuli should also be included in the model, as in Multilevel Structural Equation models (MSEM) [@Hoyle_et_al_2023, chap. 26].

@@

<!-- Brennen generalizability theory, generalize conclusions over groups. To generalize over judges, you need to estimate them. Akin to Frisch-Wough-Lovel theorem. -->

>>>>

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone's theory deals with pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp. 267]. The theory proposes that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their underlying perception of the stimulus' trait level. According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (mean), referred to as the *modal discriminal process*, indicates the stimulus' position on the trait continuum, and its dispersion, known as the *discriminal dispersion*, reflects the variability in the perceived trait level of the stimulus.

@@

Indeed, the theory assumes that the observed dichotomous outcome arises from the distribution of the difference between the underlying discriminal processes of the stimuli, known as the *discriminal difference*. Since the individual discriminal processes follow a Normal distribution, their difference also follows a Normal distribution [@Andrich_1978]. The mode (mean) of this distribution, representing the (average) relative separation, is given by the difference between the modal discriminal processes of the stimuli $S_{BA} = S_{B} - S_{A}$. Meanwhile, the dispersion of the distribution, reflecting the variability in the relative separation, is calculated as $\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}$. Here, $\sigma_{B}$ and $\sigma_{A}$ are the previously defined discriminal dispersions, while $\rho$ measures the correlation between the discriminal processes of the stimuli. This correlation quantifies the dependence of the judge's perception of the trait in one stimulus on his perception of the same trait in another.

@@

<!-- 6. example of the effects of correlation on the discriminal difference -->
@fig-correlation illustrates how varying correlations influence the distribution of the discriminal difference for the two texts depicted in @fig-discriminal_process. Since the texts differ in quality, higher correlations increase the likelihood that the discriminal difference distinctly favors text B over text A. This is evident from the larger proportion of area under the curve, $P(B > A)$, that lies above zero (shaded gray area in @fig-correlation). Conversely, it is not hard to infer that if the texts had similar or identical quality levels, higher correlations would likely reduce the chance that the discriminal difference distinctly favors one text over the other. This probability reduction occurs because the distribution of the discriminal difference would become more narrowly centered around zero (not illustrated).

@@

<!-- 1. Thurstone's Cases and their simplifying assumptions  -->
The previous section outlines the general form of Thurstone's theory, which applies to a CJ design where a single judge evaluates multiple stimuli. For the practical application of the theory, Thurstone developed four additional cases derived from this general form, where each successive case incorporates additional simplifying assumptions. Case I represents the general form of the theory. Case II extends this by allowing multiple judges to make comparisons rather than restricting the comparisons to a single judge. Case III introduces the assumption of zero correlation between stimuli. Case IV builds on this by assuming that the stimuli have similar dispersions. Finally, Case V replaces this assumption with the condition that the stimuli have equal discriminal dispersions. @tbl-thurstone_cases summarizes these cases and their assumptions. For a detailed discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248-253].

@@

<!-- 5. The first scenario where the assumption does not hold -->
In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between the stimuli due to certain judges' biases that resist cancellation. Research on text quality suggests that when judges evaluate these traits, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. These additional relevant characteristics, which are unlikely to be equally weighted or opposing, can unevenly influence judges' perceptions, creating biases in their judgments and, consequently, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp. 346]. For example, this could occur when a judge assessing the argumentative quality of a text places disproportionate emphasis on grammatical accuracy, ultimately favoring texts with fewer errors but weaker arguments. While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the idea that the factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 7. But what are the challenges when this assumption is violated? -->
In any case, the psychometric and statistical literature emphasizes the importance of addressing factors that create dependencies between stimuli, as failing to do so can lead to inaccurate conclusions about a trait's reliability and, by extension, its validity. For instance, neglecting additional traits relevant to the stimuli, such as judges' biases, often leads to a dimensional mismatch in the model. This mismatch can potentially inflate the reliability of the trait [@Hoyle_et_al_2023, pp. 341] or, worse, introduce bias into the trait's estimates [@Ackerman_1989]. Similarly, failing to account for hierarchical (grouping) structures reduces the precision of model parameter estimates, further amplifying the overestimation of reliability [@Hoyle_et_al_2023, pp. 482]. These issues collectively undermine the trait's reliability, and because "validity cannot exist without reliability" [@Perron_et_al_2015, pp. 2], they ultimately compromise the validity of results derived from the the data modeling.

@@

<!-- 8. How to solve the violation of the assumption? -->
Fortunately, the same literature provides solutions for addressing the factors that create dependencies between stimuli. @Andrich_1978 and @Wainer_et_al_1978 propose integrating judges' biases into the CJ model. This integration would produce a model similar to a multidimensional or two-parameter logistic IRT model [@Hoyle_et_al_2023, chap. 15], depending on the theoretical assumptions a researcher is willing to make regarding the judges' biases. The literature also suggests incorporating relevant hierarchical structures into the model to account for dependencies in the stimuli, resulting in a model resembling a Multilevel Structural Equation Model (MSEM) [@Hoyle_et_al_2023, chap. 26].

<!-- Brennen generalizability theory, generalize conclusions over groups. To generalize over judges, you need to estimate them. Akin to Frisch-Wough-Lovel theorem for latent variables?. -->
